The powder x-ray diffractogram open database

### 2024-04-12 paper/00_abstract.tex
New title idea

In the current landscape of powder x-ray diffraction analysis, machine learning has emerged as a very promising direction of research resolving this bottleneck by enabling fully automated powder diffraction analysis.
A noteable difficulty in applying machine learning to this domain is the lack of labeled datasets i.e.\
diffraction patterns originating from experiment with accompanying structure solutions, which has relegated machine learning
researches to training on simulated data.
Since these simulations largely fail to accurately reflect experiment, the performance of models trained
on this simulated data, also by in large fails to transfer to experiment and provide value in practice.
We aim to remedy this by providing a freely available and easily accessible dataset of partially labeled
powder diffraction data, providing machine learning researches with a large quantity of real experimental diffractograms
collected from a broad range of samples.
We hope that this can guide machine learning research towards more realistic
simulations and potentially enable training and testing models on real experimental data.

>>>>

In the current landscape of powder x-ray diffraction analysis, machine learning has emerged as a very promising direction of research resolving this bottleneck by enabling fully automated powder diffraction analysis.
A noteable difficulty in applying machine learning to this domain is the lack of labeled datasets i.e.\
diffraction patterns originating from experiment with accompanying structure solutions, which has relegated machine learning
researchers to training on simulated data.
Since these simulations largely fail to accurately reflect experiment, the performance of models trained
on this simulated data, also by in large fails to transfer to experiment and provide value in practice.
We aim to remedy this by providing a freely available and easily accessible dataset of partially labeled
powder diffraction data, providing machine learning researchers with a large quantity of real experimental diffractograms
collected from a broad range of samples.
We hope that this can guide machine learning research towards more realistic
simulations and potentially enable training and testing models on real experimental data.

### 2024-04-12 paper/01_introduction.tex
New title idea

However due to the scarcity of labeled diffractograms originating from experiment, collecting a labeled dataset
of real experimental powder x-ray diffraction data of sufficient size to train a model
that can be useful in practice is next to impossible and is not the aim of this dataset.
Rather we hope that this dataset can enable machine learning researchers to evaluate how closely their simulations
represent actual experimental data, identify what features are missing and modify their simulations algorithms
accordingly to arrive at a simulation that nearly perfectly mirrors how data appears in reality. \\

Secondly we would like to also present this dataset as a challenge: The fraction of labeled data can be used to
easily numerically evalute the performance of the model on a broad range of data as it appears in experiment, which can be used
to gauge how well models would perform in practice. \\
But even the unlabeled portion of the dataset can be used to test the performance of models: The least a model
has to do in order to prove useful in practice is to make a guess about the crystal structure that is good enough
for Rietveld refinement to converge toward the true structure. \\
Hence, the performance of the model can be tested by evaluating if the inital guess of the model is sufficient for
Rietveld refinment to refine the initial guess to a structure that produces a diffractogram matching the one observed in experiment.
Finally, by inviting the application of models to data as it appears in experiment we would like to bring attention
to some of the practical difficulties that need to be addressed in applying a model to a wide variety of x-ray diffractograms
collected from different sources: For example the angle and the step size of the diffractograms will in general be variable
which is something that any model that aims to perform well on this dataset but be equipped to handle.

@@

Machine learning techniques have emerged as a powerful tool in the toolkit of
materials scientists. While they are often used to make predictions on
the properties of materials or find new materials with certain properties, an increasingly interesting domain is the automated analysis of raw experimental measurements
guided by machine learning\supercite{radovicMachineLearningEnergy2018}.

With the advent of high-throughput experiments, the amount of gathered data is
vast and the analysis often becomes a bottleneck in the processing pipeline
\supercite{rahmanianEnablingModularAutonomous2022}. Powder X-ray diffraction
(XRD) is an important measurement technique used to obtain structural
information from polycrystalline samples\supercite{harrisContemporaryAdvancesUse2001}. 
The diffractograms are an
information-dense fingerprint of the structure of the material. However,
analyzing these diffractograms is not an easy task\supercite{holderTutorialPowderXray2019}. 
Full structure solutions and
Rietveld refinement take time and require expert knowledge, both about the
analysis technique and the materials class at hand. This is not feasible in high-throughput experiments on a larger
scale. Therefore, the question arises whether it is possible to automatically 
analyze powder diffractograms with machine learning models trained on large
amounts of data, making it possible to run inference almost instantaneously.

During the last few years, there have been several studies tackling this
objective by applying machine learning models to various tasks concerning the
analysis of powder diffractograms such as phase
classification\supercite{leeDeeplearningTechniquePhase2020,maffettoneCrystallographyCompanionAgent2021,schuetzkeEnhancingDeeplearningTraining2021,szymanskiProbabilisticDeepLearning2021,wangRapidIdentificationXray2020},
phase fraction determination\supercite{leeDatadrivenXRDAnalysis2021}, space
group
classification\supercite{parkClassificationCrystalStructure2017,oviedoFastInterpretableClassification2019,zalogaCrystalSymmetryClassification2020,vecseiNeuralNetworkBased2019,suzukiSymmetryPredictionKnowledge2020,chakrabortyDeepCrystalStructure2022},
machine-learning-guided Rietveld
refinement\supercite{ozakiAutomatedCrystalStructure2020,fengMethodArtificialIntelligence2019},
extraction of lattice
parameters\supercite{dongDeepConvolutionalNeural2021,chitturiAutomatedPredictionLattice2021,chakrabortyDeepCrystalStructure2022,
habershonPowderDiffractionIndexing2004} and crystallite
sizes\supercite{dongDeepConvolutionalNeural2021,chakrabortyDeepCrystalStructure2022},
and also novelty detection based on unsupervised techniques\supercite{bankoDeepLearningVisualization2021}.
Since an abundant source of experimental diffractograms is hard to come by, most
applications train their models on simulated diffractograms from the Inorganic
Crystal Structure Database (ICSD)\supercite{bergerhoff1987}, which contains a total of 272\,260 structures (October 2022).

\citeauthor{leeDeeplearningTechniquePhase2020} used a deep convolutional neural
network (CNN) trained on a large dataset of multiphase compositions from the
quaternary \ce{Sr-Li-Al-O} pool to classify present phases in the
diffractogram\supercite{leeDeeplearningTechniquePhase2020}. In a follow-up
study, they further showed good results for phase fraction inference in the
quaternary \ce{Li-La-Zr-O} pool\supercite{leeDatadrivenXRDAnalysis2021}.
\citeauthor{schuetzkeEnhancingDeeplearningTraining2021} performed phase
classification on iron ores and cement compounds and used data augmentation with
respect to lattice parameters, crystallite sizes, and preferred orientation
\supercite{schuetzkeEnhancingDeeplearningTraining2021}. They showed that
especially the lattice parameter variations enhance the classification accuracy
significantly.

Instead of the analysis of phase composition,
\citeauthor{dongDeepConvolutionalNeural2021} performed regression of scale
factors, lattice parameters, and crystallite sizes in a five-phase catalytic
materials system\supercite{dongDeepConvolutionalNeural2021}. In contrast to
supervised tasks, \citeauthor{bankoDeepLearningVisualization2021} used a
variational autoencoder to visualize variations in space group, preferred
orientation, crystallite size, and peak
shifts\supercite{bankoDeepLearningVisualization2021}.
\citeauthor{parkClassificationCrystalStructure2017} used a deep CNN to classify
space groups of single-phase diffractograms, reaching a test accuracy of 81.14\%
on simulated diffractograms.\supercite{parkClassificationCrystalStructure2017}
However, as we will show later in this paper, this accuracy is highly overestimated and drops to 56.1\% when test splits are designed in a way to reduce data leakage in non-IID datasets such as the ICSD.
\citeauthor{vecseiNeuralNetworkBased2019}\supercite{vecseiNeuralNetworkBased2019} developed
a similar approach and applied their classifier to experimental diffractograms
from the RRUFF mineral database\supercite{lafuentePowerDatabasesRRUFF2015},
reaching an experimental test accuracy of 54\%.

>>>>

However due to the scarcity of labeled diffractograms originating from experiment, collecting a labeled dataset
of real experimental powder x-ray diffraction data of sufficient size to train a model
that can be useful in practice is next to impossible and is not the aim of this dataset.
Rather we hope that this dataset can enable machine learning researchers to evaluate how closely their simulations
represent actual experimental data, identify what features are missing and modify their simulations algorithms
accordingly to arrive at a simulation that nearly perfectly mirrors how data appears in reality.\\
One very simple approach which could give a first hint of how closely simulations reflect experiment would be to attempt
to train a classifier on simulated and real diffractograms.
If the classifier performs well that would be an indicator that the simulation doesn't reflect experiment sufficiently well.
Note that the converse doesn't necessarily hold as well.
For a set of given sample properties there is only one corresponding simulated diffractogram, but there may an infinite amount of
variations in how the diffractogram of a sample could look like in experiment for example due varying measurement apperatus, noise
, sample impurities and lattice defects.
In order for a simulated dataset to closely resemble experiment it is critical that the simulatio also represents these
statistical variations.
Determining the nature of these variations is a much a harder task than making individual diffractograms look like they
originated from experiment and cannot be accomplished by simply training a classifier.
Nevertheless, we hope that this dataset can at least provide clues of how to take this aspect into account during
simulation as well.\\

Secondly we would like to also present this dataset as a challenge: The fraction of labeled data can be used to
easily numerically evalute the performance of the model on a broad range of data as it appears in experiment, which can be used
to gauge how well models would perform in practice. \\
But even the unlabeled portion of the dataset can be used to test the performance of models: The least a model
has to do in order to prove useful in practice is to make a guess about the crystal structure that is good enough
for Rietveld refinement to converge toward the true structure. \\
Hence, the performance of the model can be tested by evaluating if the inital guess of the model is sufficient for
Rietveld refinment to refine the initial guess to a structure that produces a diffractogram matching the one observed in experiment.
Finally, by inviting the application of models to data as it appears in experiment we would like to bring attention
to some of the practical difficulties that need to be addressed in applying a model to a wide variety of x-ray diffractograms
collected from different sources: For example the angle and the step size of the diffractograms will differ between sources
, which is something that any model that aims to perform well on this dataset must be equipped to handle.

@@

%
%
%Machine learning techniques have emerged as a powerful tool in the toolkit of
%materials scientists. While they are often used to make predictions on
%the properties of materials or find new materials with certain properties, an increasingly interesting domain is the automated analysis of raw experimental measurements
%guided by machine learning\supercite{radovicMachineLearningEnergy2018}.
%
%With the advent of high-throughput experiments, the amount of gathered data is
%vast and the analysis often becomes a bottleneck in the processing pipeline
%\supercite{rahmanianEnablingModularAutonomous2022}. Powder X-ray diffraction
%(XRD) is an important measurement technique used to obtain structural
%information from polycrystalline samples\supercite{harrisContemporaryAdvancesUse2001}.
%The diffractograms are an
%information-dense fingerprint of the structure of the material. However,
%analyzing these diffractograms is not an easy task\supercite{holderTutorialPowderXray2019}.
%Full structure solutions and
%Rietveld refinement take time and require expert knowledge, both about the
%analysis technique and the materials class at hand. This is not feasible in high-throughput experiments on a larger
%scale. Therefore, the question arises whether it is possible to automatically
%analyze powder diffractograms with machine learning models trained on large
%amounts of data, making it possible to run inference almost instantaneously.
%
%During the last few years, there have been several studies tackling this
%objective by applying machine learning models to various tasks concerning the
%analysis of powder diffractograms such as phase
%classification\supercite{leeDeeplearningTechniquePhase2020,maffettoneCrystallographyCompanionAgent2021,schuetzkeEnhancingDeeplearningTraining2021,szymanskiProbabilisticDeepLearning2021,wangRapidIdentificationXray2020},
%phase fraction determination\supercite{leeDatadrivenXRDAnalysis2021}, space
%group
%classification\supercite{parkClassificationCrystalStructure2017,oviedoFastInterpretableClassification2019,zalogaCrystalSymmetryClassification2020,vecseiNeuralNetworkBased2019,suzukiSymmetryPredictionKnowledge2020,chakrabortyDeepCrystalStructure2022},
%machine-learning-guided Rietveld
%refinement\supercite{ozakiAutomatedCrystalStructure2020,fengMethodArtificialIntelligence2019},
%extraction of lattice
%parameters\supercite{dongDeepConvolutionalNeural2021,chitturiAutomatedPredictionLattice2021,chakrabortyDeepCrystalStructure2022,
%habershonPowderDiffractionIndexing2004} and crystallite
%sizes\supercite{dongDeepConvolutionalNeural2021,chakrabortyDeepCrystalStructure2022},
%and also novelty detection based on unsupervised techniques\supercite{bankoDeepLearningVisualization2021}.
%Since an abundant source of experimental diffractograms is hard to come by, most
%applications train their models on simulated diffractograms from the Inorganic
%Crystal Structure Database (ICSD)\supercite{bergerhoff1987}, which contains a total of 272\,260 structures (October 2022).
%
%\citeauthor{leeDeeplearningTechniquePhase2020} used a deep convolutional neural
%network (CNN) trained on a large dataset of multiphase compositions from the
%quaternary \ce{Sr-Li-Al-O} pool to classify present phases in the
%diffractogram\supercite{leeDeeplearningTechniquePhase2020}. In a follow-up
%study, they further showed good results for phase fraction inference in the
%quaternary \ce{Li-La-Zr-O} pool\supercite{leeDatadrivenXRDAnalysis2021}.
%\citeauthor{schuetzkeEnhancingDeeplearningTraining2021} performed phase
%classification on iron ores and cement compounds and used data augmentation with
%respect to lattice parameters, crystallite sizes, and preferred orientation
%\supercite{schuetzkeEnhancingDeeplearningTraining2021}. They showed that
%especially the lattice parameter variations enhance the classification accuracy
%significantly.
%
%Instead of the analysis of phase composition,
%\citeauthor{dongDeepConvolutionalNeural2021} performed regression of scale
%factors, lattice parameters, and crystallite sizes in a five-phase catalytic
%materials system\supercite{dongDeepConvolutionalNeural2021}. In contrast to
%supervised tasks, \citeauthor{bankoDeepLearningVisualization2021} used a
%variational autoencoder to visualize variations in space group, preferred
%orientation, crystallite size, and peak
%shifts\supercite{bankoDeepLearningVisualization2021}.
%\citeauthor{parkClassificationCrystalStructure2017} used a deep CNN to classify
%space groups of single-phase diffractograms, reaching a test accuracy of 81.14\%
%on simulated diffractograms.\supercite{parkClassificationCrystalStructure2017}
%However, as we will show later in this paper, this accuracy is highly overestimated and drops to 56.1\% when test splits are designed in a way to reduce data leakage in non-IID datasets such as the ICSD.
%\citeauthor{vecseiNeuralNetworkBased2019}\supercite{vecseiNeuralNetworkBased2019} developed
%a similar approach and applied their classifier to experimental diffractograms
%from the RRUFF mineral database\supercite{lafuentePowerDatabasesRRUFF2015},
%reaching an experimental test accuracy of 54\%.
